---
title: "Data visualization -- FSA and rfishbase data exploration"
author: "(Your name here)"
date: "2/3/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
lab_path = dirname(rstudioapi::getSourceEditorContext()$path)
knitr::opts_knit$set(root.dir = lab_path)
getwd()
set.seed(9999)
require(FSAdata) 
require(FSA) # you may need to install this package
require(tidyverse)
require(rfishbase)
```

# FSA analyses
Basic linear and non-linear regression analyses for predicting growth patterns.

## Growth History
- Red drum age and length data from the Atlantic Coast (reported by OpenFisheries). `FL = fork length`
- Popular recreational target fish
- Lifespan of 40-45 years

```{r}
colnames(RedDrum)
data = RedDrum
glimpse(data)
dim(data) # prints number of rows, columns
```

Add x, y, and main title labels with the `labs()` function for both plot and ggplot
```{r}
plot = ggplot() + geom_point(data=data, mapping=aes(x=age, y=fl),
               colour = "black",
               size = 3,  # point size
               alpha = 0.6, # point transparency
               pch=21) +  # point shape
            labs(x = "Fork Length (cm)",
                 y = "Age (years)",
                 title = "Length Age Relationship")
plot
```

Use histograms to visualize distributions.
```{r}
RDhist = ggplot(data, aes(fl)) + geom_histogram(binwidth = 20, colour="black", fill="grey") +
          xlab("Fork Length (mm)") + ylab("Count") + ggtitle("Red Drum length distribution")
RDhist
```

```{r}
RDhist2 <- ggplot(data, aes(age)) + geom_histogram(binwidth = 3, colour="black", fill="grey") +
            xlab("Age (years)") + ylab("Count") + ggtitle("Red Drum age distribution")
RDhist2
```

Boxplots are also useful to displaying distributions.
```{r}
RDbox <- ggplot(data, aes(factor(age), fl)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  xlab("Age (years)") + ylab("Fork Length (mm)") + ggtitle("Red Drum growth distribution")
RDbox
```

## Biological Growth Patterns

`glimpse` Ruffe data from Superior Lake catches in 1992.
```{r}
ruf = RuffeSLRH92
colnames(ruf)
glimpse(ruf)
```

We can use the `sapply` function to run a for loop. Here, we will tell the function to check all columns for `NA` values. We'll get rid of empty rows for length and weight.   

```{r}
sapply(ruf,function(x) sum(is.na(x)))
na_rows = which(is.na(ruf$length) | is.na(ruf$weight))
ruf = ruf[-c(na_rows),]
```

Let's plot Length vs Weight relationship of Ruffe with `plot` and `ggplot` and add some labels. 
```{r}
ruf_plot = ggplot() + geom_point(data=ruf, mapping=aes(x=length, y=weight)) +
  labs(x = "Total Length (mm)",
       y = "Weight (g)",
       title = "Length Weight Relationship of Ruffe")
ruf_plot
```


We can transform data with the `log()` function to discover patterns in the data. Here, we show the **linear** relationship between Length and Weight. 
```{r}
ruf_plot = ggplot() + geom_point(data=ruf, mapping=aes(x=log(length), y=log(weight))) +
  labs(x = "Total Length (mm)",
       y = "Weight (g)",
       title = "Length Weight Relationship of Ruffe")
ruf_plot
```

Create a simple linear regression model with `lm()`
```{r}
lm1 <- lm(data=ruf, log(weight)~log(length))
# Basic summary information is extracted by submitting the saved lm() object to summary() as follows
summary(lm1)
```
Fit our line
```{r}
ggplot() + geom_point(data=ruf, aes(x=log(length), y=log(weight))) + 
  geom_line(data=ruf, aes(x=log(length), y=lm1$fitted.values), color="red") + 
  xlab("log Total Length (mm)") + 
  ylab("log Weight (g)")
```

```{r}
par(mfrow=c(2,2))
plot(lm1)
```

In **ggplot** we can fit the same regression model with the `geom_smooth()` or `stat_smooth()` functions.
```{r}
ggplot(ruf, aes(log(length),log(weight))) +
  geom_point() +
geom_smooth(method="lm", formula=y~x, color="red") +
  xlab("log Total Length (mm)") + 
  ylab("log Weight (g)") +
annotate("text", x=3.4, y=3, label= paste("R2 =", round(summary(lm1)$r.squared, 3)))
```

Visualize `length`, `weight` and `sex`
```{r}
ggplot() + 
  geom_point(data=ruf, aes(x=length, y=weight, color=sex)) + 
  xlab("Total Length (mm)") + 
  ylab("Weight (g)") +
  ggtitle("Length at Weight of Ruffe") + 
  scale_color_manual(values = c("female" = "maroon", "male" = "skyblue3", "unknown" = "grey55"))
```

We can also use the shape 
```{r}
ggplot() + 
  geom_point(data=ruf, aes(x=length, y=weight, shape=sex, color=sex), size=2.5) + 
  xlab("Total Length (mm)") + ylab("Weight (g)") +
  ggtitle("Length at Weight of Ruffe") + 
  scale_color_manual(values = c("female" = "maroon", "male" = "skyblue3", "unknown" = "grey55"))
```

We can use stacked barplots to do the same thing.  
```{r}
ggplot(ruf, aes(length)) +
  geom_bar(aes(fill=sex),width=3) + 
  xlab("Total Length (mm)") + 
  scale_fill_manual(values = c("female" = "maroon", "male" = "skyblue3", "unknown" = "grey55")) + 
  ggtitle("Length Distribution of Ruffe")
```

Now, let's model the `Weight-Length` relationship of Ruffe with `W=aL^b` where `W`=weight, `L`=Length, `a`=rate of change (intercept), `b`=pattern growth of species.
$$
W = aL^b
$$

A *linear* equation is in the form `y = mx + b`.
This is a non-linear equation because `x` in this case `L` is to a power. 
Use `nls()` to develop a non-linear model. Parameters `a` and `b` are estimated from the given values. Our model will estimate the best parameters by minimizing residuals.
```{r}
model_ruf = nls(weight~a*length^b, start=list(a=160,b=3), data=ruf)
summary(model_ruf)
coef(model_ruf)
```

Use `predict()` to calculate the y-value (weight) based on the model we created above
```{r}
ggplot(ruf, aes(x=length,y=weight)) +
  geom_point() +
  geom_line(data=ruf, aes(x=length,y=predict(model_ruf)), col="red")+
  labs(x = "Length (mm)",
       y = "Weight (g)",
       title = "Weight Length relationship of Lake Superior Ruffe")
```

We can evaluate our model by plotting the residuals
```{r}
plot(model_ruf)
```

Install `gridExtra` so that we can combine ggplots 
```{r, eval=F}
# install.packages('gridExtra')
```

```{r}
title="Weight Length relationship of Lake Superior Ruffe"
rufp1 = ggplot(ruf, aes(log(length),log(weight))) +
  geom_point() +
geom_line(aes(x=log(length), y=log(predict(model_ruf))), color="red") +
  labs(x = "log (Length)",
       y = "log (Width)",
       title = title)

rufp2 = ggplot(ruf, aes(x=length,y=weight)) +
  geom_point() +
  geom_line(aes(x=length,y=predict(model_ruf)),col="red")+
  labs(x = "Length (mm)",
       y = "Weight (g)",
       title = "")

gridExtra::grid.arrange(rufp1, rufp2)
```

von Bertalanffy growth function (VBGF) in R -- Red Drum
[resource](https://www.r-bloggers.com/2019/12/von-bertalanffy-growth-plots-i/)
$$
L(a) = L_{inf}(1-e^{(-k(a-t_0))})
$$

```{r}
vb = vbFuns(param="Typical")
```

```{r}
f.starts <- vbStarts(fl~age,data=RedDrum) # vbStarts is an optimization algorithm designed to find the best params
f.fit <- nls(fl~vb(age,Linf,K,t0),data=RedDrum,start=f.starts)
coef(f.fit)
```

```{r}
ggplot() + 
  geom_point(data=RedDrum, aes(x=age, y=fl)) + 
  geom_line(data=RedDrum, aes(x=age, y=predict(f.fit)), color="red") + 
  xlab("Age (years)") + 
  ylab("Fork Length (mm)") + 
  ggtitle("Red Drum Length at Age")
```

Compare size and maturation relationships of groundfish vs groundsharks.  

`popchar` Wmax, Lmax, age (tmax)
`maturity` minimum age at maturation (AgeMatMin), minimum length at maturation (LengthMatMin)
`popgrowth` size at maturity (Lm), age (tmax)

```{r, fig.height=4}
library(rfishbase)

```

