---
title: "Life History and Population Dynamics -- part 1"
author: "(Your name here)"
date: "2/2/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
lab_path = dirname(rstudioapi::getSourceEditorContext()$path)
knitr::opts_knit$set(root.dir = lab_path)
getwd()
set.seed(9999)
require(FSAdata) 
require(FSA) # you may need to install this package
require(tidyverse)
require(rfishbase)
```

# FSA analyses
Basic linear and non-linear regression analyses for predicting growth patterns.

## Growth History
### Red Drum
- Red drum age and length data from the Atlantic Coast (reported by OpenFisheries). `FL = fork length`
- Popular recreational target fish
- Lifespan of 40-45 years

```{r}
colnames(RedDrum)
data = RedDrum
glimpse(data)
dim(data) # prints number of rows, columns
```

Add x, y, and main title labels with the `labs()` function for both plot and ggplot
```{r}
plot = ggplot() + geom_point(data=data, mapping=aes(x=age, y=fl),
               colour = "black",
               size = 3,  # point size
               alpha = 0.6, # point transparency
               pch=21) +  # point shape
            labs(x = "Fork Length (cm)",
                 y = "Age (years)",
                 title = "Length at Age Relationship")
plot
```

Use histograms to visualize distributions.
```{r}
RDhist = ggplot(data, aes(fl)) + geom_histogram(binwidth = 20, colour="black", fill="grey") +
          xlab("Fork Length (mm)") + ylab("Count") + ggtitle("Red Drum Length Distribution")
RDhist
```

```{r}
RDhist2 <- ggplot(data, aes(age)) + geom_histogram(binwidth = 3, colour="black", fill="grey") +
            xlab("Age (years)") + ylab("Count") + ggtitle("Red Drum Age Distribution")
RDhist2
```

Boxplots are also useful to displaying distributions.
```{r}
RDbox <- ggplot(data, aes(factor(age), fl)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  xlab("Age (years)") + ylab("Fork Length (mm)") + ggtitle("Red Drum Growth Distribution")
RDbox
```

## Biological Growth Patterns
### Ruffe
`glimpse` Ruffe data from Superior Lake catches in 1992. 

The ruffe is an aggressive fish. It has a rapid population growth. The ruffe eats a variety of foods. The ruffe has a lack of natural predators which creates the potential to displace other species in newly invaded areas and to cause the native fish populations to decrease.
```{r}
ruf = RuffeSLRH92
colnames(ruf)
```

We can use the `sapply` function to run a for loop. Here, we will tell the function to check all columns for `NA` values. We'll get rid of empty rows for length and weight.   
```{r}
sapply(ruf,function(x) sum(is.na(x))) # ?sapply
na_rows = which(is.na(ruf$length) | is.na(ruf$weight))
ruf = ruf[-c(na_rows),]
```

Plot Length vs Weight relationship of Ruffe with `plot` and `ggplot` and add some labels. 
```{r}
ruf_plot = ggplot() + geom_point(data=ruf, mapping=aes(x=length, y=weight)) +
  labs(x = "Total Length (mm)",
       y = "Weight (g)",
       title = "Length and Weight Relationship of Ruffe")
ruf_plot
```

We can transform data with the `log()` function to discover patterns in the data. Here, we show the **linear** relationship between Length and Weight. 
```{r}
ruf_plot = ggplot() + geom_point(data=ruf, mapping=aes(x=log(length), y=log(weight))) +
  labs(x = "log Total Length (mm)",
       y = "log Weight (g)",
       title = "Length and Weight Relationship of Ruffe")
ruf_plot
```

Create a simple linear regression model with `lm()`
```{r}
lm1 <- lm(data=ruf, log(weight)~log(length))
# Basic summary information is extracted by submitting the saved lm() object to summary() as follows
summary(lm1)
```
Fit our line
```{r}
ggplot() + geom_point(data=ruf, aes(x=log(length), y=log(weight))) + 
  geom_line(data=ruf, aes(x=log(length), y=lm1$fitted.values), color="red") + 
  xlab("log Total Length (mm)") + 
  ylab("log Weight (g)")
```

Look at the goodness of fit.
```{r}
par(mfrow=c(2,2))
plot(lm1)
# Residuals - difference between observed and predicted values
# QQ plot - comparing two probability distributions 
# Scale-Location - spread of residuals
# Leverage - influence of data points on residuals
```

In **ggplot** we can fit the same regression model with the `geom_smooth()` or `stat_smooth()` functions.
```{r}
ggplot(ruf, aes(log(length),log(weight))) +
  geom_point() +
geom_smooth(method="lm", formula=y~x, color="red") +
  xlab("log Total Length (mm)") + 
  ylab("log Weight (g)") +
annotate("text", x=3.4, y=3, label= paste("R2 =", round(summary(lm1)$r.squared, 3))) # retrieve R-squared value
```

## Non-linear relationships
Visualize `length`, `weight` and `sex`
```{r}
ggplot() + 
  geom_point(data=ruf, aes(x=length, y=weight, color=sex)) + 
  xlab("Total Length (mm)") + 
  ylab("Weight (g)") +
  ggtitle("Length at Weight of Ruffe") + 
  scale_color_manual(values = c("female" = "maroon", "male" = "skyblue3", "unknown" = "grey55"))
```

We can also use the shape 
```{r}
ggplot() + 
  geom_point(data=ruf, aes(x=length, y=weight, shape=sex, color=sex), size=2.5) + 
  xlab("Total Length (mm)") + ylab("Weight (g)") +
  ggtitle("Length at Weight of Ruffe") + 
  scale_color_manual(values = c("female" = "maroon", "male" = "skyblue3", "unknown" = "grey55"))
```

We can use stacked barplots to do the same thing.  
```{r}
ggplot(ruf, aes(length)) +
  geom_bar(aes(fill=sex),width=3) + 
  xlab("Total Length (mm)") + 
  scale_fill_manual(values = c("female" = "maroon", "male" = "skyblue3", "unknown" = "grey55")) + 
  ggtitle("Length Distribution of Ruffe")
```

Now, let's model the `Weight-Length` relationship of Ruffe with `W=aL^b` where `W`=weight, `L`=Length, `a`=rate of change (intercept), `b`=pattern growth of species.
$$
W = aL^b
$$

A *linear* equation is in the form `y = mx + b`.
This is a non-linear equation because `x` in this case `L` is to a power. 
Use `nls()` to develop a non-linear model. Parameters `a` and `b` are estimated from the given values. Our model will estimate the best parameters by minimizing residuals.
```{r}
model_ruf = nls(weight~a*length^b, start=list(a=160,b=3), data=ruf)
summary(model_ruf)
coef(model_ruf)
```

Use `predict()` to calculate the y-value (weight) based on the model we created above
```{r}
ggplot(ruf, aes(x=length,y=weight)) +
  geom_point() +
  geom_line(data=ruf, aes(x=length,y=predict(model_ruf)), col="red")+
  labs(x = "Length (mm)",
       y = "Weight (g)",
       title = "Weight Length relationship of Lake Superior Ruffe")
```

We can evaluate our model by plotting the residuals
```{r}
plot(model_ruf)
```

Install `gridExtra` so that we can combine ggplots 
```{r, eval=F}
# install.packages('gridExtra')
```

```{r}
title="Weight and Length Relationship of Lake Superior Ruffe"
rufp1 = ggplot(ruf, aes(log(length),log(weight))) +
  geom_point() +
geom_line(aes(x=log(length), y=log(predict(model_ruf))), color="red") +
  labs(x = "log (Length)",
       y = "log (Width)",
       title = title)

rufp2 = ggplot(ruf, aes(x=length,y=weight)) +
  geom_point() +
  geom_line(aes(x=length,y=predict(model_ruf)),col="red")+
  labs(x = "Length (mm)",
       y = "Weight (g)",
       title = "")

gridExtra::grid.arrange(rufp1, rufp2)
```

von Bertalanffy growth function (VBGF) in R -- Red Drum
[resource](https://www.r-bloggers.com/2019/12/von-bertalanffy-growth-plots-i/)
$$
L(a) = L_{inf}(1-e^{(-k(a-t_0))})
$$

```{r}
vb = vbFuns(param="Typical") # from FSA package
```

```{r}
f.starts <- vbStarts(fl~age,data=RedDrum) # vbStarts is an optimization algorithm designed to find the best params
f.fit <- nls(fl~vb(age,Linf,K,t0),data=RedDrum,start=f.starts)
coef(f.fit)
```

```{r}
ggplot() + 
  geom_point(data=RedDrum, aes(x=age, y=fl)) + 
  geom_line(data=RedDrum, aes(x=age, y=predict(f.fit)), color="red") + 
  xlab("Age (years)") + 
  ylab("Fork Length (mm)") + 
  ggtitle("Red Drum Length at Age")
```

```{r}
plot(f.fit)
# plot(nlstools::nlsResiduals(f.fit))
# Shows autocorrelation, qq plot, and standardized residuals
```

# Population Dynamics

## Mortality rate
```{r}
# ---- Mortality rate ----------------------------

# get mortality from Then.etal.2014
getMtmax = function(lifespan){		
		M = 4.899 * lifespan^-0.916
		M
}

# get mortality using Dulvy.etal.2014's method
getMDulvy.etal.2014 = function(lifespan, agemat){
	w = (lifespan+agemat)/2 # average lifespan
	M = 1/w
	M
}
```

```{r}
Mruffe = rfishbase::maturity(species_list = c("Gymnocephalus cernua"))
Gruffe = rfishbase::popgrowth(species_list = c("Gymnocephalus cernua"))
```

```{r}
agemat = rbind(Mruffe$AgeMatMin, Mruffe$AgeMatMin2)


```

## Exponential Population Growth
$$
N_t = N_0(e^{(b-d)t})
$$
```{r examplebox2.1}
b = c(0.1,0.1,0.1) # birth rates
d = c(0.04,0.1,0.125) # death rates
N0 = c(500,500,500) # starting population number
Time = seq(0,20,0.5) # time steps
Nt = matrix(NA, nrow = length(Time), ncol = 3)
for (i in 1:3){
  Nt[,i] = N0[i]*exp((b[i]-d[i])*Time)
}
plot(Nt[,1]~Time, type = "l", ylim = range(Nt))
lines(Nt[,2]~Time, col = "red")
lines(Nt[,3]~Time, col = "blue")
```

## Logistic Population Growth
$$
N_{t+1} = N_t + rN_t(1-N_t/K) - C_t 
$$
```{r}
r = 0.5 # intrinsic growth rate
K = 1000 # carrying capacity
N0 = 50 # starting population number
Time = seq(0,25,1) # time
Nt = c()
Nt[1] = N0
Ct = 10 # catch or harvest
for (i in 1:25){
  Nt[i+1] = Nt[i] + r*Nt[i]*(1-Nt[i]/K) - Ct
}
plot(Nt~Time, type = "l", ylim = range(Nt))
```
