---
title: "Data visualization -- FSA and rfishbase data exploration"
author: "(Your name here)"
date: "2/3/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
lab_path = dirname(rstudioapi::getSourceEditorContext()$path)
knitr::opts_knit$set(root.dir = lab_path)
getwd()
set.seed(9999)
require(FSAdata) 
require(FSA)
require(tidyverse)
require(rfishbase)
```

# FSA analyses
Today, we'll be using Fisheries Stock Assessment `FSA` and `rfishbase` data to look at growth patterns (i.e., Length-Weight and Length-Age). Visualizing population dynamics is critical to fisheries management and so here, we'll look at the fundamentals for plotting in R. Then we'll look at basic linear and non-linear regression analyses that are needed to predict growth patterns.

## Plot and ggplot
- Red drum age and length data from the Atlantic Coast (reported by NOAA). `FL = fork length`
- Popular recreational target fish
- Lifespan of 40-45 years

Look at individuals that are older than 5 years old (age>5)
```{r}
colnames(RedDrum)
data = RedDrum
glimpse(RD)
dim(data) # prints number of rows, columns
```

`~` is used to indicate `y vs x` on an axis. Here, we'll plot fork length vs age

```{r}
plot(fl~age,data=data)
```

Add x, y, and main title labels with the `labs()` function for both plot and ggplot
```{r}
plot = ggplot() + geom_point(data=data, mapping=aes(x=age, y=fl),
               colour = "black",
               size = 3,
               alpha = 0.6,
               pch=21) + 
            labs(x = "Fork Length (cm)",
                 y = "Age (years)",
                 title = "Length Age Relationship")
plot
```

Use histograms to visualize distributions.
```{r}
RDhist = ggplot(data, aes(fl)) + geom_histogram(binwidth = 20, colour="black", fill="grey")
RDhist
```

```{r}
RDhist2 <- ggplot(data, aes(age)) + geom_histogram(binwidth = 3, colour="black", fill="grey")
RDhist2
```

Boxplots are also useful to displaying distributions.

```{r, fig.height=3,fig.width=4.5}
RDbox <- ggplot(data, aes(factor(age), fl)) + geom_boxplot()
RDbox
```

`glimpse` Ruffe data from Superior Lake catches in 1992.

```{r}
ruf = RuffeSLRH92
colnames(ruf)
glimpse(ruf)
```

We can use the `sapply` function to run a for loop. Here, we will tell the function to check all columns for `NA` values. We'll get rid of empty rows for length and weight.   

```{r}
sapply(ruf,function(x) sum(is.na(x)))
na_rows = which(is.na(ruf$length) | is.na(ruf$weight))
ruf = ruf[-c(na_rows),]
```

Let's plot Length vs Weight relationship of Ruffe with `plot` and `ggplot` and add some labels.

```{r}
plot(ruf$length,ruf$weight, xlab = "Total Length (mm)", ylab = "Weight (g)",
     main="Length Weight Relationship of Ruffe")

p <- ggplot()

p + geom_point(data=ruf, mapping=aes(x=length, y=weight)) +
  labs(x = "Total Length (mm)",
       y = "Weight (g)",
       title = "Length Weight Relationship of Ruffe")
```


We can transform data to look at it differently with the `log()` function. Here, we show the linear relationship between Length and Weight. 

```{r}
plot(log(ruf$length),log(ruf$weight), xlab = "log Total Length (mm)", 
     ylab = "log Weight (g)")
```

Now, let's create a simple linear regression model with `lm()`

```{r}
lm1 <- lm(log(weight)~log(length),data=ruf)
# Basic summary information is extracted by submitting the saved lm() object to summary() as follows
summary(lm1)
```
Our linear model can be **fit** to a plot with `abline()` in plot. 

```{r}
plot(log(ruf$length),log(ruf$weight), xlab = "log Total Length (mm)", 
     ylab = "log Weight (g)")
abline(lm1, col="red", lwd=2)
```

How could I write a function to calculate the R^2 value of a linear model? 

```{r}
plot(log(ruf$length),log(ruf$weight), xlab = "log Total Length (mm)", 
     ylab = "log Weight (g)")
abline(lm1, col="red", lwd=2)
rsq <- function (x, y) {
  val = cor(x, y) ^ 2
  val = round(val,3)
  val
  }
text(x=3.4, y=3, col=2, cex=0.8,
     paste("R2 =", rsq(log(ruf$weight), lm1$fitted.values)))
```

In **ggplot** we can fit the same regression model with the `geom_smooth()` or `stat_smooth()` functions.

```{r}
ggplot(ruf, aes(log(length),log(weight))) +
  geom_point() +
geom_smooth(method="lm", formula=y~x)+
annotate("text", x=3.4, y=3, label= paste("R2 =", 
                          rsq(log(ruf$weight), lm1$fitted.values)))
```

We can visualize multiple variables on the same plot. Here, we'll look at **categorical** variables such as `sex` in addition to Length vs Weight. We can use `color` or `col` to visualize this variable.

```{r}
plot(ruf$length,ruf$weight, xlab = "Total Length (mm)", 
     ylab = "Weight (g)",
     main="Length Weight of Ruffe", col=ruf$sex)
legend("topleft", c("Female", "Male", "Unknown"), pch=21,col=1:3, lty=0)
```

We can also use the shape of the point to visualize other variables, with `pch`

```{r}
plot(ruf$length,ruf$weight, xlab = "Total Length (mm)", 
     ylab = "Weight (g)",
     main="Length Weight of Ruffe",col=c(1,2,3)[ruf$sex], 
     pch = c(21, 22, 23)[ruf$sex], xlim=c(100,250))
legend("bottomright", c("Female", "Male", "Unknown"), pch=21:23,col=1:3, lty=0)
```

We can use stacked barplots to do the same thing.  

```{r}
ggplot(ruf, aes(length))+geom_bar(aes(fill=sex),width=3)
```

Sometimes, we have to play around with the plots to understand how best to visualize it. Let's continue to do some exploratory analysis on Ruffe. 

Now, let's model the weight-length relationship of Ruffe with `W=aL^b` where `W`=weight, `L`=Length, `a`=rate of change (intercept), `b`=pattern growth of species. We can use `nls()` (non-linear least squares) to utilize this model.

```{r}
# W = aL^b
model_ruf = nls(weight~a*length^b, start=list(a=160,b=3), data=ruf)
summary(model_ruf)
```

Use `predict()` to calculate the y-value (weight) based on the model we created above

```{r}
ggplot(ruf, aes(x=length,y=weight)) +
  geom_point() +
  geom_line(aes(x=length,y=predict(model_ruf)),col="blue")+
  labs(x = "Length (mm)",
       y = "Weight (g)",
       title = "Weight Length relationship of Lake Superior Ruffe")

```

For doing exploratory data analysis, it's good to create panels of different plotted relationships. Unfortunately, `ggplot` and `plot` objects cannot be combined with conventional panel generators. So, we'll use one or the other here.

With `ggplot` it is simple to combine plots in a panel with the `gridExtra` package.

```{r, eval=F}
install.packages('gridExtra')
```


```{r}
title="Weight Length relationship of Lake Superior Ruffe"
rufp1 = ggplot(ruf, aes(log(length),log(weight))) +
  geom_point() +
geom_smooth(method="lm", formula=y~x)+
annotate("text", x=3.4, y=3, label= paste("R2 =", 
                          rsq(log(ruf$weight), lm1$fitted.values)))+
  labs(x = "log (Length)",
       y = "log (Width)",
       title = title)

rufp2 = ggplot(ruf, aes(x=length,y=weight)) +
  geom_point() +
  geom_line(aes(x=length,y=predict(model_ruf)),col="blue")+
  labs(x = "Length (mm)",
       y = "Weight (g)",
       title = "")

gridExtra::grid.arrange(rufp1, rufp2)
```

Now, let's use rfishbase and do some exploratory data analysis. Let's compare size and maturation relationships of groundfish vs groundsharks.  

`popchar` Wmax, Lmax, age (tmax)
`maturity` minimum age at maturation (AgeMatMin), minimum length at maturation (LengthMatMin)
`popgrowth` size at maturity (Lm), age (tmax)

```{r, fig.height=4}
library(rfishbase)

fishdat1 = popchar(species_list = species_list(Order = "Pleuronectiformes")) 
fishdat_M1 = maturity(fishdat1$Species)
fishdat_L1 = popgrowth(fishdat1$Species)
fishdat2 = popchar(species_list = species_list(Order = "Carcharhiniformes"))
fishdat_M2 = maturity(fishdat2$Species)
fishdat_L2 = popgrowth(fishdat2$Species)

p1 = ggplot(fishdat1, aes(x=tmax,y=Lmax)) +
  geom_point()+labs(x="Age", y="Maximum length", title="Pleuronectiformes")
p2 = ggplot(fishdat2, aes(x=tmax,y=Lmax)) +
  geom_point()+labs(x="Age",y="", title="Carcharhiniformes")
p3 = ggplot(fishdat_M1, aes(x=AgeMatMin,y=LengthMatMin)) +
  geom_point()+labs(x="Minimum age at maturity", y="Minimum length at maturity", title="")
p4 = ggplot(fishdat_M2, aes(x=AgeMatMin,y=LengthMatMin)) +
  geom_point()+labs(x="Minimum age at maturity", y="", title="")
p5 = ggplot(fishdat_L1, aes(x=tmax,y=Lm)) +
  geom_point()+labs(x="Age", y="Mean size at maturity", title="")
p6 = ggplot(fishdat_L2, aes(x=tmax,y=Lm)) +
  geom_point()+labs(x="Age", y="", title="")

gridExtra::grid.arrange(p1,p2,p3,p4,p5,p6,nrow=3,ncol=2)
```

Now, let's plot life history patterns for other species found in `FSA` and `rfishbase`

***Exercise***

