---
title: "Population dynamics and Von Bertalanffy indiviudal growth "
author: "Fisheries Management and Conservation Spring 2021"
date: "2/11/21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First, we will load some data and some packages to be used later in the lab - the data is in the .Dat format - you need to change the line below to direct R to the file location on your machine.

```{r}
dat <- read.table("~/Desktop/Fisheries Management/Data/codAL.Dat",head=TRUE)
```
You will need the following packages installed and loaded. 
```{r message=FALSE}
library(nlstools)
library(tidyverse)
library(magrittr)
library(FSA)
```  
Leave that data aside for now, first we will look at population growth models.

### Exponential Growth model

```{r examplebox2.1}
b = c(0.1,0.1,0.1) # birth rates
d = c(0.04,0.1,0.125) # death rates
N0 = c(500,500,500) # starting populations

Time = seq(0,20,0.5) # time steps
Nt = matrix(NA, nrow = length(Time), ncol = 3)
for (i in 1:3){
  Nt[,i] = N0[i]*exp((b[i]-d[i])*Time)
}

plot(Nt[,1]~Time, type = "l", ylim = range(Nt))
lines(Nt[,2]~Time, col = "red")
lines(Nt[,3]~Time, col = "blue")
```

**EXERCISE**  
1) Change the values for b and d and run the models again. *Think about:* How does this change the plot? Why?  
2) Plot the linear versions of these models.

### Logistic Population Growth

```{r}
r = 0.5
K = 1000
N0 = 50
Time = seq(0,25,1)
Nt = c()
Nt[1] = N0
Ct = 10
for (i in 1:25){
  Nt[i+1] = Nt[i] + r*Nt[i]*(1-Nt[i]/K) - Ct
}
par(mfrow  = c(1,2))

plot(Nt~Time, type = "l", ylim = range(Nt))

Ct = 20
for (i in 1:25){
  Nt[i+1] = Nt[i] + r*Nt[i]*(1-Nt[i]/K) - Ct
}
plot(Nt~Time, type = "l", ylim = range(Nt))



```
  
**On Your Own**  
Manipulate the parameters k, r, and N0; how does this affect the models?

**EXERCISE**  
In a new R script, write and plot an exponential growth model for a stock with initial population biomass of 1500 (hundreds of tons), an annual death rate of 0.07, and an annual harvest (catch) of 100 (hundreds of tons). Compare the effects of differential birth rates b=0.15, b=0.2, and b=0.25 on the population trajectory. Project the population growth over 30 years.  
  
### Von Bertalanffy Individual Growth

We will use the dataframe `dat` which we loaded at the start of the lab. First, we will remove all negatives or missing values for length
```{r}
index1<-which(dat$length>0)
dat <-dat[index1,]
```

#### Parameter starting values  
The Von Bertalanffy growth equation includes the parameters Linf, K, and t0. We will define estimates of these parameters to be used in a function.  As a general rule for marine fish, estimates of K=0.3 and t0=0 are a good starting point, and we can supply the maximum length in our dataset as an estimate of mean maximum length.

```{r}
parameters <- list(Linf=max(dat$length), K=0.3, t0=0)
```
#### Writing a function  
Next, we will write a function for the Von Bertalanffy growth equation.
```{r}
VBgrowth <- function(age, Linf, K, t0) Linf*(1-exp(-K*(age-t0)))
```
We can test our function by estimating length at age 3.
```{r}
VBgrowth(3, Linf=max(dat$length), K=0.3, t0=0)
```
The package `FSA` contains a function vbFuns() which we can use to slightly modify our growth function - this allows us to supply all of the parameters as one argument.
```{r}
VBgrowth<-vbFuns()
VBgrowth(3, Linf=max(dat$length), K=0.3, t0=0)
VBgrowth(3, Linf=c(max(dat$length), 0.3, 0))
```
Both of the above code lines produce the same result, but in the second, the three parameters are provided in one argument.  
  
#### Parameter estimates  
To generate estimates of our parameters, we will fit the model using the non-linear least squares method. We need to supply our function, our dataset, and our starting estimates for parameter values (which we defined above in the list `parameters`.)
```{r}
VBfit <- nls(length~VBgrowth(age, Linf, K, t0), data=dat,start=parameters)
```
To extract our parameter estimates from the list `VBfit`, we will use the `coef()` function; similarly, to get confidence intervals for each, we can use the `confint()` function.
```{r}
coef(VBfit)
confint(VBfit)
```

#### Bootstrapping  
We now have estimates based on our data, but to make our estimates more robust, we can use a technique called bootstrapping. Bootstrapping is a technique whereby the data is sampled with replacement repeatedly, and parameters are estimated from each bootstrap sample. For non-linear models (such as VBGF), we can use the function `nlsBoot()` from the package `nlstools`.
```{r}
bootFit <- nlsBoot(VBfit)
head(bootFit$coefboot)
```
We can also estimate confidence intervals from our bootstrap samples and plot them.
```{r}
confint(bootFit,plot=TRUE)
```

#### Predictions  
We can now use our updated parameter estimates to make predictions for length-at-age.
```{r}
predicts <- data.frame(age=c(1:25)) #ages for predictions
predicts$length<-predict(VBfit,predicts) #generate predicted length at age
```
The above predictions are based on our initial (non-bootstrapped) parameter estimates. We can also generate predictions based on the bootstrap samples (note that the `vbFuns()` function we used earlier assigns "t" as the variable for age). We will use the bootstrap data to project mean length at age 5, using the `apply()` function. `apply()` allows us to apply functions to vectors or lists; here we use the argument `MARGIN=1` to apply the function across rows, rather than columns, because each row contains a set of parameter estimates. We then use the `quantile()` function to generate a 95% confidence interval.
```{r}
predicts.age5 <- apply(bootFit$coefboot, MARGIN=1, FUN=VBgrowth, t=5)
quantile(predicts.age5, c(0.025,0.975))
```

#### Plotting  
Let's plot our data.
```{r}
x <- seq(0,25,length.out=500) #ages for predictions
predLength <- VBgrowth(x,Linf=coef(VBfit)) #predicted length at age
xlmts <- range(c(x,dat$age)) #Because we are plotting outside of our age data
ylmts <- range(c(predLength,dat$length)) 
plot(length~age, data=dat, xlab="Age", ylab="Total length (cm)",
     xlim=xlmts, ylim=ylmts,pch=19,col=rgb(0,0,0,1/3))
lines(predLength~x,lwd=2)
```
  
Now we can add some confidence intervals - first we need to define the dimensions.
```{r}
LCI <- UCI <- numeric(length(x))
```
Now we are going to use a `for` loop to predict confidence interbals, again using the `apply()` function.
```{r}
for (i in 1:length(x)) {
  tmp <- apply(bootFit$coefboot,MARGIN = 1,FUN=VBgrowth, t=x[i])
  LCI[i] <- quantile(tmp,0.025)
  UCI[i] <- quantile(tmp,0.975)
}
```
We'll now update our y-axis limits to include the range of the confidence intervals, and then plot again.
```{r}
ylmts <- range(c(predLength,LCI,UCI,dat$length))
plot(length~age, data=dat, xlab="Age", ylab="Total length (cm)",
     xlim=xlmts, ylim=ylmts,pch=19,col=rgb(0,0,0,1/3))
lines(predLength~x,lwd=2)
lines(UCI~x,lwd=2,lty=2, col="red")
lines(LCI~x,lwd=2,lty=2, col="red")
```

#### Check model fit  
We will check our model fit using the `residPlot()` function.
```{r}
residPlot(VBfit)
```
  
#### Log-Transform  
As you can see, as age increases, the variability of length increases (multiplicative error). If we take the natural log of each side, then we can transform the equation, fit it again using least-squares, and check the new fit.
```{r}
dat %<>% mutate(log_length=log(length))
VBLogFit <- nls(log_length~log(VBgrowth(age,Linf,K,t0)),data=dat,start=parameters)
residPlot(VBLogFit)
```

#### Exercise  
Now, open a new R script, and using the `bluecatfish.csv` file on Canvas, fit a VBGF model and generate parameter estimates, and plot the data with confidence intervals generated from bootstrapping the data.

If you complete this, install the package `FSAdata`, and try these analyses on additional datasets. `FSAdata` contains length at age data for the following species (if the package is loaded, you can access the data easily, see example).
```{r eval = FALSE}
library(FSAdata)
data <- BlueCatfish # this is for blue catfish, which we've already done
```
Other species, identified by named data frame:  
- Bonito (Australian bonito)  
- BullTroutRML2 (Bull Trout from Rocky Mountain Lakes)  
- CreekChub  
- Cabezon (Oregon)  
- Croaker2 (Atlantic Croaker)  
- DarterOnt (Channel Darters)  
- EuroPerchTH (European Perch)  
- FWDrumLE1 (Lake Erie Freshwater Drum)  
- LJCisco (Longjaw Cisco)  
- RedDrum (Atlantic Red Drum)  
- RockBassLO1 (Lake Ontario Rock Bass)  
- SardineChile  
- SpotVA2  
- StripedBass2  
- TroutBR (Brown and Rainbow Trout)  